---
title: "DeepRHealth::medcode Module Demo"
subtitle: "R/medicine 2025 Prototype Showcase" 
author: "Zhixia Ren"
date: 05/25/2025
format: 
  # revealjs:
  #   theme: sky 
  #   slide-number: true 
  #   controls: true 
  #   progress: true 
  #   
  html:
    toc: true        
    toc_float: true   
    code-fold: false  
    embed-resources: true
lang: en
editor: visual
---

# Introduction

-   Goal: Showcase the core functionalities of the `DeepRHealth::medcode` module.

-   Supports major medical coding systems (e.g., ICD, ATC) natively in R.

-   Built with inspiration from the Python-based `PyHealth.medcode`.

<!-- -->

-   **Key Features to Demo:**
    -   Code Lookup

    -   Hierarchy Navigation (Ancestors/Descendants)

    -   Cross-System Mapping

    -   ATC Specific Utilities

# Environment Setup

**Loading Package & Dependencies**

-   We'll load the package directly from source using `devtools::load_all()`.
-   This requires running the presentation from the package's root directory.
-   We also load other necessary packages like `knitr`.

```{r setup-environment, echo=TRUE, message=FALSE, warning=FALSE}
# Load required libraries for loading and display
library(devtools)
library(knitr)
library(rappdirs) # Used by DeepRHealth internally for caching
library(fs) # Used by DeepRHealth internally for caching

# Load all functions, data, etc. from the DeepRHealth package source
# IMPORTANT: Ensure R's working directory is the package root!
load_all(".")

# Optional: Display where medcode data would be cached
# cache_dir <- user_cache_dir("DeepRHealth", "medcode")
# print(paste("Data cache directory:", cache_dir))

# NOTE: Assume necessary data (e.g., ICD9CM.csv) is pre-cached for demo speed.
```

# Data Handling

## Download & Caching

The package automatically downloads required medical code datasets (CSVs) and caches them locally for faster and offline use.

-   Uses `rappdirs` to find a user-specific cache directory.
-   The `download_medcode()` function checks the cache first; only downloads if the file is missing.

**Let's see where the data is/will be cached:**

```{r show_cache_path, echo=TRUE, message=FALSE, warning=FALSE}
# Get the platform-specific cache directory path
cache_dir <- rappdirs::user_cache_dir("DeepRHealth", "medcode")
print(paste("Data cache directory:", cache_dir))
```

### **Demonstrating `download_medcode()`:**

(This step ensures the data file exists locally, downloading only if necessary)

```{r download_medcode, echo=TRUE, message=FALSE, warning=FALSE}
# Specify the dataset name (e.g., "ICD9CM")
dataset_name <- "ICD9CM"

# Call the function - returns path. Downloads ONLY if not cached.
# NOTE: For a live demo, ensure network access OR pre-cache the file!
#       We assume it's pre-cached here.
file_path <- download_medcode(name = dataset_name)

print(paste("Path for", dataset_name, ":", file_path))
print(paste("File exists:", fs::file_exists(file_path)))
```

## Loading Data (`load_medcode`)

The `load_medcode()` function is used internally by most other functions (`lookup_code`, `get_ancestors`, etc.).

1.  It first calls `download_medcode()` to ensure the data file is available locally.

2.  Then, it reads the CSV file into an R data frame (tibble) using `readr::read_csv`.

**Example: Loading the ICD9CM data**

```{r load_medcode, echo=TRUE, message=FALSE, warning=FALSE}
# Load the data using the function
# This implicitly calls download_medcode first
icd9_data_example <- load_medcode("ICD9CM")

# Confirm data is loaded by checking dimensions and showing first few rows
print(paste("Loaded ICD9CM - Dimensions:", paste(dim(icd9_data_example), collapse = " x ")))
print("First few rows:")
kable(head(icd9_data_example))
```

# Feature 1: Code Lookup (`lookup_code`)

This function retrieves the description and potentially other details for a specific medical code within a given coding system.

**Example: Look up ICD-9-CM code "428.0"**

```{r lookup-demo, echo=TRUE, message=FALSE, warning=FALSE}
# Input: code string and system name
code_info <- lookup_code(code = "428.0", system = "ICD9CM")

# Output: A tibble/data frame row with information
# Using kable() for potentially nicer table output in the presentation
kable(code_info)
```

# Feature 2: Hierarchy - Ancestors (`get_ancestors`)

This function navigates the code hierarchy *upwards* to find all parent and ancestor codes.

**Example: Find ancestors for ICD-9-CM code "428.22"** (Systolic heart failure, acute on chronic)

```{r get_ancestors-demo, echo=TRUE, message=FALSE, warning=FALSE}
# Input: code string and system name
ancestors <- get_ancestors(code = "428.22", system = "ICD9CM")

# Output: A character vector of ancestor codes
print(ancestors)
```

# Feature 3: Hierarchy - Descendants (`get_descendants`)

This function navigates the code hierarchy *downwards* to find all child and descendant codes.

**Example: Find descendants for ICD-9-CM code "428"** (Heart failure)

```{r get_descendants, echo=TRUE, message=FALSE, warning=FALSE}
# Input: code string and system name
descendants <- get_descendants(code = "428", system = "ICD9CM")

# Output: A character vector of descendant codes (can be long!)
print(head(descendants)) # Show only the first few for brevity
print(paste("Total descendants found:", length(descendants)))
```

# Feature 4: Cross-System Mapping

## Feature 4a: Supported Map

let's see which mappings (crosswalks) are currently supported: The `supported_cross()` function returns a list of identifiers for all available code system mappings within the package.

```{r supported-crosswalks-demo, echo=TRUE, message=FALSE, warning=FALSE}
# Call the function to get the list of supported crosswalks
available_crosswalks <- supported_cross()

# Print the available mapping identifiers
print(available_crosswalks)
```

## Feature 4b：Translate codes(`map_code`)

This function translates codes from one coding system to another using pre-defined mapping tables.

**Example: Map ICD-9-CM "428.0" to the CCSCM system** (Clinical Classifications Software)

```{r map_code-demo, echo=TRUE, message=FALSE, warning=FALSE}
# Input: code, source system (from), target system (to)
mapped_code <- map_code(code = "428.0", from = "ICD9CM", to = "CCSCM")

# Output: The corresponding code(s) in the target system
print(mapped_code)
```

# Feature 5: ATC Specific Functions

Now let's look at utilities specifically designed for the ATC (Anatomical Therapeutic Chemical) classification system for drugs.

## Feature 5a: ATC Level Conversion (`atc_convert`)

This utility function truncates an ATC code to get its representation at different classification levels (L1 to L5).

**Example: Convert ATC code "L01BA01" (Methotrexate)**

```{r atc_convert-demo, echo=TRUE, message=FALSE, warning=FALSE}
atc_code <- "L01BA01"

# Get code representations at different levels
print(paste("L1 (Anatomical Main Group):", atc_convert(atc_code, level = 1))) # L
print(paste("L3 (Therapeutic Subgroup):", atc_convert(atc_code, level = 3))) # L01
print(paste("L4 (Chemical Subgroup):", atc_convert(atc_code, level = 4))) # L01B
# Level 5 is usually the full substance code
```

## Feature 5b: ATC Drug-Drug Interactions (`get_ddi`)

This function loads a predefined dataset of potential Drug-Drug Interactions (DDIs), typically represented by pairs of interacting ATC codes.

**Example: Load and view the first few DDI pairs**

```{r get_ddi-demo, echo=TRUE, message=FALSE, warning=FALSE}
# Load the DDI dataset bundled with the package/data
ddi_data <- get_ddi()

# Display the structure (first few rows)
# Assumes columns are ATC_i, ATC_j
kable(head(ddi_data))

# Optionally, show total number of interactions listed
# print(paste("Total DDI records loaded:", nrow(ddi_data)))
```

# Summary & Future Work

## Summary

-   `medcode` enables standardized handling of clinical code vocabularies
-   Unified functions for code lookup, hierarchy navigation, and cross-system mapping
-   Easily extendable to support new coding systems or custom vocabularies
-   Ready for integration into R-based medical informatics pipelines

## Next step/Future work

-   **Finalize Vignettes:** Complete detailed tutorials (vignettes) showcasing common use cases and workflows.
-   **Complete `pkgdown` Website:** Finish the documentation website for easy access to all documentation (including existing help pages) and examples.
-   **Expand Coverage:** Add support for more coding systems and mapping tables based on user needs.
-   **Implement Testing:** Create comprehensive unit tests using `testthat` to ensure code quality and stability.
